A Type System for Structured Aliasing in Parallel Programs

### 2012-10-29 doc/types-pldi2013/example.tex
More edits to sections 2 3 and 4

In addition to privileges, functions can also specify {\em coherence}
on regions.  Coherence specifies what updates the function may
see from other functions using the same region.  If not otherwise
specified, coherence defaults to {\em exclusive}, meaning the
function must appear to execute in program order.  Line 14 shows
an example of {\tt atomic}, a relaxed coherence mode requiring
that operations to {\tt rn} and {\tt rg} appear atomic relative
to other functions using those regions.  The most relaxed coherence
mode is {\tt simult}; simultaneous coherence allows concurrent access 
to the region by all functions that are using the region
in a simultaneous mode.  The interaction between {\em tasks} (which are functions chosen by the runtime to execute in parallel) using the same
region with different coherence
modes is formalized in Section~\ref{sec:coherence}.

>>>>

In addition to privileges, functions can also specify {\em coherence}
on regions.  Coherence specifies what updates the function may
observe from other functions using aliased regions.  
If not otherwise specified, coherence defaults to {\em exclusive}, 
meaning the function must appear to execute in program order.  
Line 14 in Listing~\ref{lst:circuit_leaf} shows
an example of {\tt atomic}, a relaxed coherence mode requiring
that operations to {\tt rn} and {\tt rg} appear atomic relative
to other functions using regions which may alias.  The most relaxed coherence
mode is {\tt simult}; simultaneous coherence allows concurrent access 
to the region by all functions that are using the region
in a simultaneous mode.  The interaction between {\em tasks} (which are functions chosen to execute in parallel) using the same
region with different coherence
modes is formalized in Section~\ref{sec:coherence}.

### 2012-10-29 doc/types-pldi2013/langdef.tex
More edits to sections 2 3 and 4

As Legion attempts to benefit from both static and dynamic checks
to achieve performance while maintaining safety, both the type system and the operational
semantics have been extended in interesting ways from a basic expression language.  We
will describe these extensions in the next two subsections, and explain how these extensions
were chosen to preserve a straightforward proof of the soundness of the type system.

@@

The introduction of constraints into $\Omega$ occurs through the use of the {\em partition} expression,
which makes use of a new type specific to Legion, a {\em coloring}.  A coloring is an opaque mapping from
pointers (which much point into the region being colored) to integer ``color'' values.  (For simplicity, 
Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
an empty map and a {\em color} expression which adds a new entry to an existing map, but the opacity of the
coloring allows for more efficient generation (and storage) of colorings as needed.)  The {\em partition} 
expression introduces names for subregions corresponding to each of those ``colors'', guaranteeing that each
subregion is disjoint from the other subregions and all are included in the original region.

@@

Finally, the task call expression handles the aliasing of logical region names between the caller and callee and
verifies the property that the privileges held by a called task are always a subset of those held
by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
enable the hierarchical scheduling model used by the Legion runtime.

@@

The final element of the environment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
programs can
be scheduled in a hierarchical (and therefore scalable) fashion, we require a formulation of operational semantics
that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
reductions are applied and in cases of relaxed coherence), it is possible for the heap accesses performed by
one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
a global view of all
partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
step of execution.  This approach has the benefit of being constructive but the implicit global ``scheduler'' is
difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.

Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
information about the overall environment.  The information consists of the initial state of the heap $S$, and a
set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
and is used to relax the constraints the evaluation of an expression.  Whereas the requirement for locations not
in the clobber set is the usual one of {\em consistency} (i.e. a read from a location observes the effect of all 
previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithms
used by the Legion runtime satisfy these constraints, guaranteeing the soundness of the Legion programming model.

The complete operational semantics rules can be found in \cite{LegionTypes12} - we show the most interesting ones
in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
traces are constructed and how the result of a read expression is relaxed when the location falls within the
clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.
(A formal definition of the {\em valid\_interleave} predicate can be found in Figure~\ref{fig:validinterleave}, and
will be discussed in Section~\ref{sec:coherence}.)

@@

The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
operational semantics rules for the {\em newcolor} and {\em color} expressions.  The color expression accepts an
existing coloring and returns one in which the specified location is given the specified color, replacing an
existing coloring for that location (if present).  The behavior of the newcolor expression is a bit trickier.
Since Legion supports the {\em partition-then-allocate} model in which the {\em new} expression may be called
on subregions, each subregion that results from a partitioning may need an arbitrary number of locations for which
the program does not know the name at partitioning time.  The semantics for the newcolor expression allow this while
guaranteeing that any explicit colorings performed by the program are honored.

@@

Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
correct type and does so within the regions of memory for which privileges are held.  In fact, this results holds
even if the {\em valid\_interleave} constraint is weakened to just require an arbitrary interleaving of the memory
traces of subexpressions (i.e. the {\em any\_interleave} predicated defined in Figure~\ref{fig:opsecfns}).  The 
additional constraints on {\em valid\_interleave} in Figure~\ref{fig:validinterleave} address the coherence of 
heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion.

>>>>

To achieve high performance in the presence of aliasing, the Legion type system has 
been designed to efficiently leverage both static and dynamic checks.
Consequently, both the type system and the operational
semantics have been extended in interesting ways from a basic expression language.  We
describe these extensions in the next two subsections, and explain how they 
were chosen to preserve a straightforward proof of soundness for the type system.

@@

The introduction of constraints into $\Omega$ occurs through the use of the {\em partition} expression,
which makes use of a new type specific to Legion, a {\em coloring}.  A coloring is an opaque mapping from
pointers (which must point into the region being colored) to integer {\em color} values.  (For simplicity, 
Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
an empty map, and a {\em color} expression which adds a new entry to an existing map. The opacity of the
coloring allows for more efficient generation (and storage) of colorings as needed.)  The {\em partition} 
expression introduces names for subregions corresponding to each color in the coloring, guaranteeing that each
subregion is disjoint from the other subregions and all are included in the original region.

@@

Finally, the task call expression handles the aliasing of logical region names between the caller and callee and
verifies the functional requirement that the privileges held by a called task are always a subset of those held
by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
enable the hierarchical scheduling model used by the Legion runtime.

@@

The final element of the environment is the {\em clobber set} $C$.  Our goal is to show that parallel Legion
programs can
be scheduled in a hierarchical fashion.  We therefore require a formulation of operational semantics
that is both parallel and hierarchical.  Because Legion allows constrained nondeterminism (in the order in which
reductions are applied and in cases of relaxed coherence with aliasing), it is possible for the heap accesses performed by
one expression's evaluation to observe interference from other concurrent evaluations.  The traditional way of
handling this in operational semantics \cite{XYZ} is to use a top-down approach with small-step semantics and
a global view of all
partially-evaluated expressions.  Rules are specified for which expression(s) are valid choices for the next small
step of execution.  This approach has the benefit of being constructive but the implicit global scheduler is
difficult to map into a hierarchical one in which decisions must necessarily be made with limited information.

Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
information about the overall environment.  The information consists of the initial state of the heap $S$, and a
set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
and is used to relax the constraints the evaluation of an expression.  Whereas the requirement for locations not
in the clobber set is the usual one of {\em consistency} (i.e. a read from a location observes the effect of all 
previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.

The complete operational semantics rules can be found in \cite{LegionTypes12}; we show the most interesting ones
in Figure~\ref{fig:semantics}.  The rules for the {\em read}, {\em write}, and {\em reduce} show how dynamic memory
traces are constructed and how the result of a read expression is relaxed when the location falls within the
clobber set.  Additionally, the write and reduce operations, with their two subexpressions, demonstrate how the
traces of subexpressions are interleaved within the environment (and clobber set) of the parent expression.
(A formal definition of the {\em valid\_interleave} predicate can be found in Figure~\ref{fig:validinterleave}, and
will be discussed in Section~\ref{sec:coherence}.)

@@

The general form of a coloring being an abitrary mapping of locations to ``colors'' (integers) is shown in the
operational semantics rules for the {\em newcolor} and {\em color} expressions.  The color expression accepts an
existing coloring and returns one in which the specified location is given the specified color, replacing an
existing coloring for that location (if present).  The behavior of the newcolor expression is more subtle.
Since Legion supports the {\em partition-then-allocate} model (see Section~\ref{sec:example}) in which the {\em new} expression may be called
on subregions, each subregion that results from a partitioning may need an arbitrary number of locations for which
the program does not know the name at partitioning time.  The semantics for the newcolor expression allow this while
guaranteeing that any explicit colorings performed by the program are honored.

@@

Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
correct type and does so within the regions of memory for which privileges are held.  In fact, this results holds
even if the {\em valid\_interleave} constraint is weakened to only require an arbitrary interleaving of the memory
traces of subexpressions (i.e. the {\em any\_interleave} predicated defined in Figure~\ref{fig:opsecfns}).  The 
additional constraints on {\em valid\_interleave} in Figure~\ref{fig:validinterleave} address the coherence of 
heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion to enable parallelism
in the presence of aliasing.
